paper_draft

### 2022-09-01 paper_draft.qmd
first draft'

@@

-   the name of the organism studied

-   the particular aspect or system studied

-   the variable(s) manipulated.

@@

passive voice

(typically 150-300 words)

-   A brief introduction to the topic that you're investigating.

-   Explanation of why the topic is important in your field/s.

-   Statement about what the gap is in the research.

-   Your research question/s / aim/s.

-   An indication of your research methods and approach.

-   Your key message.

-   A summary of your key findings.

-   An explanation of why your findings and key message contribute to the field/s

@@

@@

@@

Next the Kaplan-Meier estimator was calculated, stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature. A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.

@@

The function of this section is to summarize general trends in the data without comment, bias, or interpretation. The results of statistical tests applied to your data are reported in this section although conclusions about your original hypotheses are saved for the Discussion section.

@@

The first dimension ($\lambda_! = 0.19$) accounts for 19.29% of the variance in the data 
with the first 4 dimensions of the MCA analysis accouting for 54% of the variance.

@@

The contribution of the categories can be ranked within each of the first four
MCA dimensions.

The first dimension is dominated by highly specific facial dysmophisms such as 
upward sweep of hair, priminent lower lip & bulbous nose. Also significant are cardiomyopathy and mild vermian hypoplasia. This principal dimension also notably 
is categorized by the absence of simplified gyral pattern and abnormal EEG 
with or without seizure.

The second MCA dimension also has highly contributory facial features such as short philtrum, hairline related dysmorphisms and spastic tatraparesis. This dimension also has a notable absence of typical features such as microcephaly, arthrogryposis and hypomyelination.

The third dimension is categorised by feeding and respiratory dysfunction with 
vocal cord palsy, tracheostomy, tube feeding and GERD. The fourth dimension is best described by small for gestational age and structural brain anomalies. See @fig-ctr

```{r ctr-plot, fig.height=10, fig.width=10, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 20 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(20, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  labs(y = '')
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
fviz_mca_biplot(mca,  
               repel = T) 
```

@@

```{r}
# CLValidation table or plot 
summary(cluster_select)
```

```{r}
cluster_select@measures |> 
  as_tibble(rownames = 'measure') |> 
  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
  mutate(cluster = as.numeric(str_remove(cluster, ".hierarchical"))) |> 
  ggplot(aes(cluster, metric, colour = measure, group = measure)) +
  geom_line()
```

```{r, fig.height=12, fig.width=10}
# AGNES 
plot(clustering, which.plot = 2)
```

```{r, fig.height=8, fig.width=8,  warning=FALSE}
# plot MCA axes  
cbind(cluster_data, clust = cluster_assignments) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust) |> 
  ggplot(aes(dim_1, dim_2, colour = clust)) +
  geom_point(position = "jitter") +
  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20)

```{r}
# Profile table 
cbind(cluster_data, clust = cluster_assignments) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  select(clust, var) |> 
  arrange(clust) |> 
  kable()
```

@@

```{r}
# KM Curve
x <- survminer::surv_summary(kapm_baseline, data = survival_data)
kable(x, digits = 2)
```

```{r, fig.height=8, fig.width=8,  warning=FALSE}
survminer::ggsurvplot_df(x,
                         conf.int = TRUE,
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw())

@@

@@

# box plot of variant 
ggplot(survival_data, aes(variant, survival_time)) +
  geom_boxplot()
```

```{r}
x2 <- survminer::surv_summary(kapm, data = survival_data)
kable(x2, digits = 2)
```

```{r, fig.height=8, fig.width=8,  warning=FALSE}
survminer::ggsurvplot_df(x2,
                              conf.int = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw())

```{r}
# log rank
tar_load(logrank)
logrank

```{r}
# cox
tar_load(cox_ph)
cox_ph
```

@@

```{r}
# table with med gen 
# cox power analysis
# cluster for outlier
# cluster for partitioning
```

@@

-   an integration of your results with those of previous studies in order to arrive at explanations for the observed phenomena

-   possible explanations for unexpected results and observations, phrased as hypotheses that can be tested by realistic experimental procedures, which you should describe

-   End the Discussion with a summary of the principal points you want the reader to remember

>>>>

```{r}
# table with med gen 
# Better tables with gtsummary
```

@@

Statistical Clinical Phenotype and Patient Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA)

@@

A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with in children with bi-allelic loss-of-function variants in *SMPD4*.

Few case reports have been described to date, with a diverse and severe clinical phenotype in most reported cases, the progression of this condition is not well understood.

A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of indivudals with this condition. This is driven in part to the paucity of analysis-ready data on subjects.

This research aims to collate and standardise available case reports, to analyse and identify
meaningful clusters in the clinical phenotype and the quantify the survival probability
function for children with NEDMABA.

To overcome the challenge of sparse, multidimensional data on very few subjects, we 
employ Multiple Correspondece Analysis as a dimension reduction technique, which is 
then subject to heirarchical clustering and interpretation. To quantify more accurate 
survival probabilities, kaplan-meier estimation is formulated to account for censoring
in the data.

Several distinct clusters of clinical features were discovered in the current case reports
which can help explain the variation present in the clinical phenotype for children 
with this condition. Furthermore, survival probability for those affected declines
sharply in early infancy, but exhibits a wide range of outcomes provisionally 
associated with variant type.

We are able to identify the classic phenotype for this condition, as well as two
other auxillary feature clusters. The first relates findings of vocal cord paralysis and
swallowing dysfunction, the other relates to more complex brain anomalies. In addition, 
several outlying or non-typical features are able to be classified. Despite strong
indications of differences in survival outcomes based on variant type, interpretation
of these results are guarded based on very low sample sizes.

Ultra rare conditions are a challenge for clinicians and families, with often limited
information on diverse outcomes. Despite challenges with sparse and inconsistent data, 
analytical techniques to better describe associations and trajectories of indivudals
with these conditions can provide clinicians and genetic counsellors with 
enhanced information to aide in decision making and support for families.

@@

\[NEED TO INCLUDE CONTRIBUTION\]

@@

\[insert clvalid details\]

@@

Next the Kaplan-Meier estimator was calculated, \[ON WHAT?\] stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature. A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.

@@

@@

The initial results of the MCA were inspected with a bi-plot and identified two individuals (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse. These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and wer subject to a missense variation in SMPD4. Further analysis of this influence is conducted below. For now, these subjects were removed to allow for more meaningful analysis.

Repeating MCA for the remaining subjects shows the first dimension ($\lambda_! = 0.18$) accounts for 17.88% of the variance in the data with the first 4 dimensions of the MCA analysis accounting for 54% of the variance.

```{r}
tar_load(mca_prelim)
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr.

The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline. Also significant are seizure and developmental delay.

The second MCA dimension is categorised by feeding and respiratory dysfunction with vocal cord palsy, tracheostomy, tube feeding and GERD.

Examining the bi-plot of the first two principal axes @fig-biplot shows contrasting sources of variation between conditions relating to airway and swallowing function and predominantly structural brain anomalies such as as cerebellar hypoplasia, agenesis of corpus collosum and development delay.

The third dimension is characterised by complex non-typical features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features. These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.

```{r ctr-plot, fig.height=10, fig.width=10, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 20 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  labs(y = '')
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

Internal cluster validation measures resulted in between 2 and 6 clusters as the optimal choice.

Proceeding with the optimal cluster selection, per the Silhouette coefficient of 6 clusters.

A scatterplot of the cluster results @fig-clust projected on just the first two dimensions highlights a core cluster of classic features of NEDMABA. A separate cluster is formed of airway and feeding related conditions involing vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles.

A cluster relating to structual brain anomalies is articulated with conditions such as cerebellar hypoplasia and agenesis of corpus collosum, along with small for gestational age.

Some smaller satellite clusters exists where highly distinctive features were associated with an indivual or family. Also a singleton clusters exist for features that are distinct or unusual such as posteriorly angulated ears.

```{r}
cluster_select@measures |> 
  as_tibble(rownames = 'measure') |> 
  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
  mutate(`optimal cluster no.` = as.numeric(str_remove(cluster, ".hierarchical")))  |> 
  select(-cluster) |> 
  group_by(measure) |> 
  filter(metric == min(metric)) |> 
  kable(digits = 2, caption = "Optimal cluster numbers determined by four similar metrics measuring cluster analysis performance.")
```

```{r, fig.width=8,  warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."
cluster_select@measures |> 
  as_tibble(rownames = 'measure') |> 
  pivot_longer(cols =  -measure, names_to = 'cluster', values_to = "metric") |> 
  mutate(cluster = as.numeric(str_remove(cluster, ".hierarchical"))) |> 
  ggplot(aes(cluster, metric, colour = measure, group = measure)) +
  geom_line() +
  labs(x = 'number of clusters') +
  facet_wrap(~measure, scales= 'free')
```

```{r, fig.height=12, fig.width=10}
#| label: fig-dendro
#| fig-cap: "Dendrogram of heirarchical clustering method"
plot(clustering, which.plots = 2, main = NA, sub = NA, xlab = NA)
```

```{r, fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with cluster assignment indicated by colour."
cbind(cluster_data, clust = cluster_assignments) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust) |> 
  ggplot(aes(dim_1, dim_2, colour = clust)) +
    geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20) +
  geom_point(aes(colour = clust), position = "jitter") +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       colour = 'cluster')
```

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with survival probability at time 0 of 0.82 (0.69, 0.98). Survival probability after one year at 0.403 (0.252, 0.644) with median survival of 150 days and survival probability at close to ten years of 0.24 (0.11, 0.52).

```{r}
# KM Curve
x <- survminer::surv_summary(kapm_baseline, data = survival_data)
kable(x, digits = 2, caption = "Survival probabilities derived from Kaplan-Meier estimation. ")
```

```{r, fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-surv
#| fig-cap: "Kaplan Meier survival curve"
survminer::ggsurvplot_df(x,
                         color = "black",
                         conf.int = TRUE,
                         conf.int.fill = "gray",
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(),
                         legend = "none") +
  labs(x = "time (days)")

@@

A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations. However, even within missense hymozygous variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as indeed noted by @magini2019loss.

@@

# box plot of variant 
ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
  geom_boxplot()
```

The log rank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.

```{r}
# log rank
tar_load(logrank)
logrank
```

@@

@@

The Multiple Correspondence Analysis confirmed the heterogeneous variation in the data with the first four princicple components being required to explain more than 50% of the variation in the dataset.

The examination of key contributing factors, as well as a visual inspection of the dat projected into this lower dimensional space, indicate the primary sources of variation in the clinical phenotype can be explained by few indivudals with very different clinical features. This has confounded the typical or classical presentation.

Beyond these outliers, a interesting disticntion is made between the hallmark features of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay with specific airway and swallowing features such as vocal cord paralysis, tracheostomy, GERD and tube feeding.

Further distictions are possible with more specific structural brainanomalies exhibiting as distinct sources of variation in the clininal phenotype.

Many of the outlying features identifes are specific facial or anatomical dysmophims. While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail.

On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination. However, there was significant heterogenity in how feet contractures were reported in clinical case reports. A distinction was made between reported cases with rocker bottom feed or congenital vertical talus and those with clubfeet/talipes. In practice its not clear how these diagnoses were arived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literaure where possible.

The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves. This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child. While the number of clusters chosen all represent valid partition of the data, a larger number was selected to align the goals of the analysis which was to uncover distinct groups. A smaller number would have likely yeilded isolated outliers from the main body of the features. This still occurs, however more substructure is detected with the higher cluster number.

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into the childhood and in some cases into the second decade of life. The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.

Despite NEDMABA exhibting a diverse and hetergeneous clinical phenotype, it is possible to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis. This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay. Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD. In additional to typical MRI findings of symplified gyral pattern and hypomyelination, there exists clusters of structural anomalies of the corpus calllosum and cereballa hypoplasia.

The survival of indivuduals with this condition is critical given early demise in infancy and childhood is a noted feature. This work identifies the survival probabilites and shows a one year survival rate of 0.403 (0.252, 0.644). A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57. This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to overestimate the survival probability.

## Appendix

```{r}
16/28
# Profile table 
cbind(cluster_data, clust = cluster_assignments) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  select(clust, var) |> 
  arrange(clust) |> 
  kable()
```
